\begin{thebibliography}{99}

% Ouvrages académiques fondamentaux
\bibitem{versteeg2007introduction}
Versteeg, H. K., \& Malalasekera, W. (2007). \emph{An Introduction to Computational Fluid Dynamics: The Finite Volume Method} (2nd ed.). Pearson Education.

\bibitem{heller2011scale}
Heller, V. (2011). Scale effects in physical hydraulic engineering models. \emph{Journal of Hydraulic Research}, 49(3), 293-306.

\bibitem{ferziger2002computational}
Ferziger, J. H., \& Perić, M. (2002). \emph{Computational Methods for Fluid Dynamics} (3rd ed.). Springer.

\bibitem{roychowdhury2020computational}
Roychowdhury, D. G. (2020). \emph{Computational Fluid Dynamics for Incompressible Flows}.

\bibitem{patankar1980numerical}
Patankar, S. V. (1980). \emph{Numerical Heat Transfer and Fluid Flow}. Hemisphere Publishing Corporation.

\bibitem{tu2018computational}
Tu, J., Yeoh, G. H., \& Liu, C. (2018). \emph{Computational Fluid Dynamics: A Practical Approach} (3rd ed.). Butterworth-Heinemann.

\bibitem{schlichting2016boundary}
Schlichting, H., \& Gersten, K. (2016). \emph{Boundary-Layer Theory} (9th ed.). Springer.

\bibitem{anderson1995computational}
Anderson, J. D. (1995). \emph{Computational Fluid Dynamics: The Basics with Applications}. McGraw-Hill.

\bibitem{batchelor1967introduction}
Batchelor, G. K. (1967). \emph{An Introduction to Fluid Dynamics}. Cambridge University Press.

\bibitem{panton2013incompressible}
Panton, R. L. (2013). \emph{Incompressible Flow} (4th ed.). John Wiley \& Sons.

\bibitem{currie2012fundamental}
Currie, I. G. (2012). \emph{Fundamental Mechanics of Fluids} (4th ed.). CRC Press.

\bibitem{white2015fluid}
White, F. M. (2015). \emph{Fluid Mechanics} (8th ed.). McGraw-Hill Education.

\bibitem{kundu2015fluid}
Kundu, P. K., Cohen, I. M., \& Dowling, D. R. (2015). \emph{Fluid Mechanics} (6th ed.). Academic Press.

\bibitem{wendt2009computational}
Wendt, J. F. (2009). \emph{Computational Fluid Dynamics: An Introduction} (3rd ed.). Springer.

\bibitem{tennekes1972first}
Tennekes, H., \& Lumley, J. L. (1972). \emph{A First Course in Turbulence}. MIT Press.

\bibitem{pope2000turbulent}
Pope, S. B. (2000). \emph{Turbulent Flows}. Cambridge University Press.

\bibitem{davidson2015turbulence}
Davidson, P. A. (2015). \emph{Turbulence: An Introduction for Scientists and Engineers} (2nd ed.). Oxford University Press.

\bibitem{reynolds1883experimental}
Reynolds, O. (1883). An experimental investigation of the circumstances which determine whether the motion of water shall be direct or sinuous. \emph{Philosophical Transactions of the Royal Society}, 174, 935-982.

\bibitem{wilcox2006turbulence}
Wilcox, D. C. (2006). \emph{Turbulence Modeling for CFD} (3rd ed.). DCW Industries.

\bibitem{sagaut2006large}
Sagaut, P. (2006). \emph{Large Eddy Simulation for Incompressible Flows: An Introduction} (3rd ed.). Springer.

\bibitem{moin1998direct}
Moin, P., \& Mahesh, K. (1998). Direct numerical simulation: a tool in turbulence research. \emph{Annual Review of Fluid Mechanics}, 30(1), 539-578.

\bibitem{lesieur2005large}
Lesieur, M., Métais, O., \& Comte, P. (2005). \emph{Large-Eddy Simulations of Turbulence}. Cambridge University Press.

\bibitem{piomelli1999large}
Piomelli, U. (1999). Large-eddy simulation: achievements and challenges. \emph{Progress in Aerospace Sciences}, 35(4), 335-362.

\bibitem{reynolds1895dynamical}
Reynolds, O. (1895). On the dynamical theory of incompressible viscous fluids and the determination of the criterion. \emph{Philosophical Transactions of the Royal Society of London A}, 186, 123-164.

\bibitem{hinze1975turbulence}
Hinze, J. O. (1975). \emph{Turbulence} (2nd ed.). McGraw-Hill.

\bibitem{rodi1993turbulence}
Rodi, W. (1993). \emph{Turbulence Models and Their Application in Hydraulics} (3rd ed.). CRC Press.

\bibitem{durbin2010statistical}
Durbin, P. A., \& Pettersson Reif, B. A. (2010). \emph{Statistical Theory and Modeling for Turbulent Flows} (2nd ed.). John Wiley \& Sons.

\bibitem{launder1972lectures}
Launder, B. E., \& Spalding, D. B. (1972). \emph{Lectures in Mathematical Models of Turbulence}. Academic Press.

\bibitem{boussinesq1877essai}
Boussinesq, J. (1877). Essai sur la théorie des eaux courantes. \emph{Mémoires présentés par divers savants à l'Académie des Sciences}, 23(1), 1-680.

\bibitem{schmitt2007boussinesq}
Schmitt, F. G. (2007). About Boussinesq's turbulent viscosity hypothesis. \emph{Comptes Rendus Mécanique}, 335(9-10), 617-628.

\bibitem{prandtl1925bericht}
Prandtl, L. (1925). Bericht über Untersuchungen zur ausgebildeten Turbulenz. \emph{Zeitschrift für Angewandte Mathematik und Mechanik}, 5(2), 136-139.

\bibitem{prandtl1945uber}
Prandtl, L. (1945). Über ein neues Formelsystem für die ausgebildete Turbulenz. \emph{Nachrichten von der Akademie der Wissenschaften in Göttingen}, 6-19.

\bibitem{hanjalic2011modelling}
Hanjalić, K., \& Launder, B. E. (2011). \emph{Modelling Turbulence in Engineering and the Environment: Second-Moment Routes to Closure}. Cambridge University Press.

\bibitem{jones1972prediction}
Jones, W. P., \& Launder, B. E. (1972). The prediction of laminarization with a two-equation model of turbulence. \emph{International Journal of Heat and Mass Transfer}, 15(2), 301-314.

\bibitem{launder1974application}
Launder, B. E., \& Sharma, B. I. (1974). Application of the energy-dissipation model of turbulence to the calculation of flow near a spinning disc. \emph{Letters in Heat and Mass Transfer}, 1(2), 131-138.

\bibitem{launder1974numerical}
Launder, B. E., \& Spalding, D. B. (1974). The numerical computation of turbulent flows. \emph{Computer Methods in Applied Mechanics and Engineering}, 3(2), 269-289.

\bibitem{wilcox1988reassessment}
Wilcox, D. C. (1988). Reassessment of the scale-determining equation for advanced turbulence models. \emph{AIAA Journal}, 26(11), 1299-1310.

\bibitem{wilcox2008formulation}
Wilcox, D. C. (2008). Formulation of the k-ω turbulence model revisited. \emph{AIAA Journal}, 46(11), 2823-2838.

\bibitem{menter1994two}
Menter, F. R. (1994). Two-equation eddy-viscosity turbulence models for engineering applications. \emph{AIAA Journal}, 32(8), 1598-1605.

\bibitem{menter2003ten}
Menter, F. R., Kuntz, M., \& Langtry, R. (2003). Ten years of industrial experience with the SST turbulence model. \emph{Turbulence, Heat and Mass Transfer 4}, 625-632.

\bibitem{hirsch2007numerical}
Hirsch, C. (2007). \emph{Numerical Computation of Internal and External Flows} (2nd ed.). Butterworth-Heinemann.

\bibitem{tannehill1997computational}
Tannehill, J. C., Anderson, D. A., \& Pletcher, R. H. (1997). \emph{Computational Fluid Mechanics and Heat Transfer} (2nd ed.). Taylor \& Francis.

\bibitem{smith1985numerical}
Smith, G. D. (1985). \emph{Numerical Solution of Partial Differential Equations: Finite Difference Methods} (3rd ed.). Oxford University Press.

\bibitem{strikwerda2004finite}
Strikwerda, J. C. (2004). \emph{Finite Difference Schemes and Partial Differential Equations} (2nd ed.). SIAM.

\bibitem{zienkiewicz2013finite}
Zienkiewicz, O. C., Taylor, R. L., \& Zhu, J. Z. (2013). \emph{The Finite Element Method: Its Basis and Fundamentals} (7th ed.). Butterworth-Heinemann.

\bibitem{reddy2006introduction}
Reddy, J. N. (2006). \emph{An Introduction to the Finite Element Method} (3rd ed.). McGraw-Hill.

\bibitem{hughes2012finite}
Hughes, T. J. R. (2012). \emph{The Finite Element Method: Linear Static and Dynamic Finite Element Analysis}. Dover Publications.

\bibitem{leveque2002finite}
LeVeque, R. J. (2002). \emph{Finite Volume Methods for Hyperbolic Problems}. Cambridge University Press.

\bibitem{eymard2000finite}
Eymard, R., Gallouët, T., \& Herbin, R. (2000). Finite volume methods. \emph{Handbook of Numerical Analysis}, 7, 713-1018.

\bibitem{moukalled2016finite}
Moukalled, F., Mangani, L., \& Darwish, M. (2016). \emph{The Finite Volume Method in Computational Fluid Dynamics: An Advanced Introduction with OpenFOAM® and Matlab®}. Springer.

\bibitem{butcher2016numerical}
Butcher, J. C. (2016). \emph{Numerical Methods for Ordinary Differential Equations} (3rd ed.). John Wiley \& Sons.

\bibitem{wesseling2001principles}
Wesseling, P. (2001). \emph{Principles of Computational Fluid Dynamics}. Springer.

\bibitem{leonard1979stable}
Leonard, B. P. (1979). A stable and accurate convective modelling procedure based on quadratic upstream interpolation. \emph{Computer Methods in Applied Mechanics and Engineering}, 19(1), 59-98.

\bibitem{leonard1991ultimate}
Leonard, B. P. (1991). The ULTIMATE conservative difference scheme applied to unsteady one-dimensional advection. \emph{Computer Methods in Applied Mechanics and Engineering}, 88(1), 17-74.

\bibitem{thompson1999handbook}
Thompson, J. F., Soni, B. K., \& Weatherill, N. P. (1999). \emph{Handbook of Grid Generation}. CRC Press.

\bibitem{knupp1993fundamentals}
Knupp, P. M., \& Steinberg, S. (1993). \emph{Fundamentals of Grid Generation}. CRC Press.

\bibitem{roache1998verification}
Roache, P. J. (1998). \emph{Verification and Validation in Computational Science and Engineering}. Hermosa Publishers.

\bibitem{schroeder2006visualization}
Schroeder, W., Martin, K., \& Lorensen, B. (2006). \emph{The Visualization Toolkit: An Object-Oriented Approach to 3D Graphics} (4th ed.). Kitware.

\bibitem{hansen2011visualization}
Hansen, C. D., \& Johnson, C. R. (2011). \emph{Visualization Handbook}. Academic Press.

\bibitem{openfoam2022guide}
OpenFOAM Foundation. (2022). \emph{OpenFOAM User Guide}. The OpenFOAM Foundation.

\bibitem{greenshields2020openfoam}
Greenshields, C. J. (2020). \emph{OpenFOAM User Guide}. The OpenFOAM Foundation.

\bibitem{jasak1996error}
Jasak, H. (1996). \emph{Error Analysis and Estimation in the Finite Volume Method with Applications to Fluid Flows}. PhD Thesis, Imperial College London.

\bibitem{rusche2002computational}
Rusche, H. (2002). \emph{Computational Fluid Dynamics of Dispersed Two-Phase Flows at High Phase Fractions}. PhD Thesis, Imperial College London.

\bibitem{patankar1972calculation}
Patankar, S. V., \& Spalding, D. B. (1972). A calculation procedure for heat, mass and momentum transfer in three-dimensional parabolic flows. \emph{International Journal of Heat and Mass Transfer}, 15(10), 1787-1806.

\bibitem{caretto1973two}
Caretto, L. S., Gosman, A. D., Patankar, S. V., \& Spalding, D. B. (1973). Two calculation procedures for steady, three-dimensional flows with recirculation. \emph{Proceedings of the Third International Conference on Numerical Methods in Fluid Mechanics}, 60-68.

\bibitem{issa1986solution}
Issa, R. I. (1986). Solution of the implicitly discretised fluid flow equations by operator-splitting. \emph{Journal of Computational Physics}, 62(1), 40-65.

\bibitem{holzmann2017mathematics}
Holzmann, T. (2017). \emph{Mathematics, Numerics, Derivations and OpenFOAM®}. Holzmann CFD.

\bibitem{courant1928uber}
Courant, R., Friedrichs, K., \& Lewy, H. (1928). Über die partiellen Differenzengleichungen der mathematischen Physik. \emph{Mathematische Annalen}, 100(1), 32-74.

\bibitem{leveque2007finite}
LeVeque, R. J. (2007). \emph{Finite Difference Methods for Ordinary and Partial Differential Equations}. SIAM.

\bibitem{trefethen1996finite}
Trefethen, L. N. (1996). \emph{Finite Difference and Spectral Methods for Ordinary and Partial Differential Equations}. Cornell University.

\bibitem{greenshields2020constant}
Greenshields, C. J. (2020). \emph{OpenFOAM User Guide - Constant Directory}. The OpenFOAM Foundation.

\bibitem{openfoam2022boundary}
OpenFOAM Foundation. (2022). \emph{OpenFOAM v10 User Guide - Boundary Conditions}. The OpenFOAM Foundation.

\bibitem{ahrens2005paraview}
Ahrens, J., Geveci, B., \& Law, C. (2005). ParaView: An end-user tool for large-data visualization. \emph{The Visualization Handbook}, 717-731.

\bibitem{ayachit2015paraview}
Ayachit, U. (2015). \emph{The ParaView Guide: A Parallel Visualization Application}. Kitware.

\bibitem{hirt1981volume}
Hirt, C. W., \& Nichols, B. D. (1981). Volume of fluid (VOF) method for the dynamics of free boundaries. \emph{Journal of Computational Physics}, 39(1), 201-225.

\bibitem{rider1998reconstructing}
Rider, W. J., \& Kothe, D. B. (1998). Reconstructing volume tracking. \emph{Journal of Computational Physics}, 141(2), 112-152.

\bibitem{scardovelli1999direct}
Scardovelli, R., \& Zaleski, S. (1999). Direct numerical simulation of free-surface and interfacial flow. \emph{Annual Review of Fluid Mechanics}, 31(1), 567-603.

% Références professionnelles et techniques (documents du barrage)
\bibitem{scet1994ahl}
SCET MAROC. (1994). \emph{Étude préliminaire du barrage Ait M'zal}. Direction des Aménagements Hydrauliques.

\bibitem{fiche_synoptique}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Fiche synoptique}.

\bibitem{fiche_historique}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Fiche historique}.

\bibitem{donnees_hydrologiques}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Données hydrologiques}.

\bibitem{note_geologique_2003}
Agence du Bassin Hydraulique du Souss Massa. (2003). \emph{Barrage Ait M'zal - Note de synthèse géologique et géotechnique n°5}.

\bibitem{fiche_organes_restitution}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Fiche organes de restitution}.

\bibitem{bilans_apports}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Bilans, fournitures, apports, bathymétrie}.

\bibitem{note_calcul_evacuateur}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Note de calcul de l'évacuateur de crues}.

\bibitem{note_calcul_vidange}
Agence du Bassin Hydraulique du Souss Massa. (s.d.). \emph{Barrage Ahl Souss (Ait M'zal) - Note de calcul de la vidange de fond}.

% Références sur les logiciels
\bibitem{galland1991telemac}
Galland, J. C., Goutal, N., \& Hervouet, J. M. (1991). TELEMAC: A new numerical model for solving shallow water equations. \emph{Advances in Water Resources}, 14(3), 138-148.

\bibitem{hervouet2007hydrodynamics}
Hervouet, J. M. (2007). \emph{Hydrodynamics of Free Surface Flows: Modelling with the Finite Element Method}. John Wiley \& Sons.

\bibitem{flow2016user}
Flow Science. (2016). \emph{FLOW-3D User Manual, Version 11.2}. Flow Science, Inc.

\bibitem{ansys2020fluent}
ANSYS. (2020). \emph{ANSYS Fluent User's Guide}. ANSYS Inc.

\bibitem{matsson2022introduction}
Matsson, J. E. (2022). \emph{An Introduction to ANSYS Fluent 2022}. SDC Publications.

\bibitem{chen1998lattice}
Chen, S., \& Doolen, G. D. (1998). Lattice Boltzmann method for fluid flows. \emph{Annual Review of Fluid Mechanics}, 30(1), 329-364.

\bibitem{kruger2017lattice}
Krüger, T., Kusumaatmaja, H., Kuzmin, A., Shardt, O., Silva, G., \& Viggen, E. M. (2017). \emph{The Lattice Boltzmann Method: Principles and Practice}. Springer.

\bibitem{cd2019star}
CD-adapco. (2019). \emph{STAR-CCM+ User Guide}. Siemens PLM Software.

\bibitem{peric2005advantage}
Perić, M., \& Ferguson, S. (2005). The advantage of polyhedral meshes. \emph{Technical Report, CD-adapco}.

\bibitem{weller1998tensorial}
Weller, H. G., Tabor, G., Jasak, H., \& Fureby, C. (1998). A tensorial approach to computational continuum mechanics using object-oriented techniques. \emph{Computers in Physics}, 12(6), 620-631.

\bibitem{jasak2009openfoam}
Jasak, H. (2009). OpenFOAM: Open source CFD in research and industry. \emph{International Journal of Naval Architecture and Ocean Engineering}, 1(2), 89-94.

\bibitem{simscale2021documentation}
SimScale GmbH. (2021). \emph{SimScale Documentation}. SimScale GmbH.

\bibitem{bletzinger2017cloud}
Bletzinger, K. U., \& Wüchner, R. (2017). Cloud-based simulation in engineering. \emph{Proceedings of the 7th GACM Colloquium on Computational Mechanics}.

\bibitem{dhi2020mike}
DHI. (2020). \emph{MIKE 21 \& MIKE 3 Flow Model FM - User Guide}. DHI Group.

\bibitem{comsol2021reference}
COMSOL. (2021). \emph{COMSOL Multiphysics Reference Manual}. COMSOL AB.

\bibitem{autodesk2021revit}
Autodesk. (2021). \emph{Autodesk Revit User Guide}. Autodesk Inc.

\bibitem{eastman2011bim}
Eastman, C., Teicholz, P., Sacks, R., \& Liston, K. (2011). \emph{BIM Handbook: A Guide to Building Information Modeling for Owners, Managers, Designers, Engineers and Contractors} (2nd ed.). John Wiley \& Sons.

\bibitem{kensek2014building}
Kensek, K., \& Noble, D. (2014). \emph{Building Information Modeling: BIM in Current and Future Practice}. John Wiley \& Sons.

\bibitem{szalapaj2005contemporary}
Szalapaj, P. (2005). \emph{Contemporary Architecture and the Digital Design Process}. Architectural Press.

\bibitem{krygiel2008green}
Krygiel, E., \& Nies, B. (2008). \emph{Green BIM: Successful Sustainable Design with Building Information Modeling}. Sybex.

\end{thebibliography}
